\pagebreak
\newpage
\TOCadd{Abstract}

\noindent \textbf{Supervisory Committee}
\tpbreak
\panel

\begin{center}
\textbf{ABSTRACT}
\end{center}
Modelling deformable tissues has applications in many interesting areas of research
including visual effects in the movies, games and surgical simulation systems. Modelling
such objects requires mastering several complex software packages. Designers
often take an incremental fine-tuning approach, i.e. a process which is based on many
small modifications to the original design. The result of such changes are required
to be seen in real-time. Creating life-like animations is far more complicated and
requires certain skills and experience.

In a surgical simulation system for example the ability to interact with the models
requires modelling the physical properties of the materials,
accurate and high performance rendering. Cutting and poking the models
also impose many challenges. Prototyping models for surgical scenes requires many
post-processing steps in order to bring those models to life.
In this research we propose a comprehensive framework for high-performance rendering
and physically-based animation of deformable  tissues using implicit surfaces.
Our system provides interactive cutting ability using smooth intersection surfaces.
Complex models can be created with implicit primitives, blending operators, affine
transformations, deformations and constructive solid geometry in a design environment that organizes
all these in a scene graph data structure called the BlobTree. We show that the
BlobTree modelling approach provides a very compact data structure which supports
incremental changes and network-based cooperative design.
Using a finite element approach we discretize BlobTree models with a GPU-Assisted
algorithm for accurate physically-based animation. Interactions with the model are
supported through smooth progressive cut surfaces. We show an application of our
system in a human skull craniotomy simulation.


\pagebreak


